# Missingness

## Introduction

When fitting models to a particular data set, missingness is of particular 
concern. Missing data can lead to biased coefficient estimates. Therefore, a 
good understanding of the different missingness patterns, their implications on
model fitting, and how to deal with them is essential. There are three 
missingness patterns (missingness mechanisms):

* missingness completely at random (MCAR)
* missingness at random (MAR)
* missingness not at random (MNAR).

In the following sections, we will use a simulated data set containing data on 
patients and their weights to illustrate these different mechanisms.

## Missingess Patterns

### Simulation Setup

Our simulation setup involves simulating the weight of patients visiting their
general physician (GP) for a checkup. For each patient, the sex (male, female)
is always recorded. The data set is balanced in terms of sex, and we assume a
normally distributed weight (kg) with a mean weight $\mu_{f} = 65$ for women and
$\mu_{m} = 78$ for men. The standard deviation $\sigma = 10$ is the same for
both sexes:

$$
W_{f} \sim \mathcal{N}(65, 100)
$$

and

$$
W_{m} \sim \mathcal{N}(78, 100).
$$

We also simulate the BMI for these patients. We assume the BMI is normally
distributed as follows:

$$
BMI \sim \mathcal{N}(25, 9).
$$

```{r}
# Simulate patient data. The data set contains 10000 observations, and is 
# balanced in terms of sex.
set.seed(123)

n <- 10000

# Define the parameters that define the distribution of weight and BMI
sd_bmi <- 3
mu_bmi <- 25

sd_weight <- 10
mu_weight_f <- 65
mu_weight_m <- 78

weights <- tibble::tibble(
  idx = 1:n,
  sex = rep(x = c("F", "M"), each = n / 2),
  weight = c(
    rnorm(n / 2, mu_weight_f, sd_weight),
    rnorm(n / 2, mu_weight_m, sd_weight)
  ),
  bmi = rnorm(n, mu_bmi, sd_bmi)
)
```

We can see that the conditional distributions of weight given sex have different 
means.

```{r echo = FALSE}
ggplot(weights, aes(x = weight)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ sex, nrow = 2, ncol = 1) +
  theme_bw() +
  labs(
    title = "Distribution of Weight Conditional on Sex",
    x = "Weight (kg)", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

But the conditional distributions of BMI are the same.

```{r echo = FALSE}
ggplot(weights, aes(x = bmi)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ sex, nrow = 2, ncol = 1) +
  theme_bw() +
  labs(
    title = "Distribution of BMI Conditional on Sex",
    x = "BMI ", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Missingness Completely at Random

Missingness completely at random indicates that the observations with missing
values are a random subset of all observations. There are no factors that affect
the probability of missingness, nor is the probability of missingess affected by
the outcome itself (weight or BMI). The distribution of observed and missing
values is the same.

```{r}
# Simulate MCAR: assume that 30% of observations have a missing outcome
set.seed(123)
idxs <- sample(x = n, size = n * 0.3)
weights["mcar"] <- weights$idx %in% idxs
```

Because the observations with missing values are a random subset of all
observations, we should see no difference between the distributions of missing
and observed values.

```{r echo = FALSE}
ggplot(weights, aes(x = weight)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ mcar, nrow = 2, ncol = 1) +
  theme_bw() +
  labs(
    title = "Distribution of Weight Conditional on Missingness (MCAR)",
    x = "Weight (kg)", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Missingness at Random

With missingness at random, we believe there are certain factors that affect the
probability of missingness, and that these same factors also have an effect on
the outcome. In our example, we see that sex affects the weight. We now assume
that sex also affects the probability of missingness.

```{r}
# Determine the conditional probabilities of missingness
p_missing_f <- 0.45
p_missing_m <- 0.25

idxs <- c(
  sample(
    x = weights[weights$sex == "F", "idx", drop = T],
    size = n * p_missing_f
  ),
  sample(
    x = weights[weights$sex == "M", "idx", drop = T],
    size = n * p_missing_m
  )
)
weights["mar"] <- weights$idx %in% idxs
```

Inspection of the distributions of observed and missing values shows us that the
distribution of missing values is shifted to the left compared to the
distribution of observed values.

```{r echo = FALSE}
ggplot(weights, aes(x = weight)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ mar, nrow = 2, ncol = 1) +
  theme_bw() +
  labs(
    title = "Distribution of Weight Conditional on Missingness (MAR)",
    x = "Weight (kg)", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

Because the probability of missingness is higher for women then for men, the
fraction of patients with a missing weight that are female is higher than the
fraction of patients with a missing weight that are men. This affects the
distribution of the missing weights, i.e., shifting the distribution to the
left. Assume that sex is the only factor that affects the probability of
missingness, then, conditional on sex, the distribution of observed values and
missing values is the same.

```{r echo = FALSE}
ggplot(weights, aes(x = weight)) +
  geom_histogram(bins = 50) +
  facet_grid(rows = vars(mar), cols = vars(sex)) +
  theme_bw() +
  labs(
    title = "Distribution of Weight Conditional on Sex and Missingness (MAR)",
    x = "Weight (kg)", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

We can see that, conditional on sex, the distribution of observed and missing
weights is the same. This illustrates the concept of missingness at random. Sex
is the factor that influences both the probability of missingness and the
outcome itself. Conditioning on this factors removes the difference in
distributions of observed and missing weights.

In practice, it is impossible to conclude missingness at random by comparing the
distribution of observed and missing values as the missing values are, of
course, missing. This example just demonstrates the concept of missingness at
random.

### Missingness Not at Random

With missingness not at random, the probability of missingness is not affected 
by a factor that is recorded, but is affected by factors that are not recorded 
or even by the outcome itself. For this missingness pattern, we will not focus
on the weight of the patient but on the patient's BMI. The distribution of BMI 
is the same for women and for men:

```{r echo = FALSE}
ggplot(weights, aes(x = bmi)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ sex, nrow = 2, ncol = 1) +
  theme_bw() +
  labs(
    title = "Distribution of BMI Conditional on Sex",
    x = "BMI", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

However, we could argue the following: if a patient visits the GP, then the GP
will more likely weigh the patient if he/she seems overweight or obese. We could
make the same argument for underweight patients, but for the sake of
illustration, we will assume the likelihood of being weighted is higher for
overweight and obese patients. Based on BMI, we can classify patients as:

* underweight: $BMI < 18.5$
* normal: $18.5 \leq BMI \leq 24.9$
* overweight: $25 \leq BMI \leq 29.9$
* obese: $30 < BMI$.

We assume the following conditional probabilities of missingness:

```{r}
# Define conditional probabilities of missingness
p_missing_uw <- 0.95 # underweight
p_missing_nw <- 0.85 # normal weight
p_missing_ow <- 0.75 # overweight
p_missing_ob <- 0.65 # obese
```

```{r}
# Classify patients based on their BMI
weights <- weights %>%
  dplyr::mutate(
    bmi_class = dplyr::case_when(
      bmi <= 18.5 ~ "uw",
      dplyr::between(bmi, 18.5, 25.0) ~ "nw",
      dplyr::between(bmi, 25.0, 29.9) ~ "ow",
      bmi >= 29.9 ~ "ob"
    )
  )
```

```{r}
fd_bmi_classes_df <- weights %>%
  dplyr::group_by(bmi_class) %>%
  dplyr::summarize(n = n())

fd_bmi_classes <- fd_bmi_classes_df$n
names(fd_bmi_classes) <- fd_bmi_classes_df$bmi_class
```

```{r}
set.seed(123)

idxs <- c(
  sample(
    x = weights[weights$bmi_class == "uw",, drop = F]$idx,
    size = round(fd_bmi_classes["uw"] * p_missing_uw)
  ),
  sample(
    x = weights[weights$bmi_class == "nw",, drop = F]$idx,
    size = round(fd_bmi_classes["nw"] * p_missing_nw)
  ),
  sample(
    x = weights[weights$bmi_class == "ow",, drop = F]$idx,
    size = round(fd_bmi_classes["ow"] * p_missing_ow)
  ),
  sample(
    x = weights[weights$bmi_class == "ob",, drop = F]$idx,
    size = round(fd_bmi_classes["ob"] * p_missing_ob)
  )
)
weights["mnar"] <- weights$idx %in% idxs
```

First, let's have a look at the distribution of BMI conditional on missingness.

```{r echo = FALSE}
ggplot(weights, aes(x = bmi)) +
  geom_histogram(bins = 50) +
  facet_wrap(~ mnar, nrow = 2, ncol = 1) +
  theme_bw() +
  labs(
    title = "Distribution of BMI Conditional on Missingness (MNAR)",
    x = "BMI", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

The distribution of the missing BMI values is shifted to the left compared to 
the distribution of the observed BMI values. This makes sense as the probability
of the weight being recorded, and thus the BMI being calculated, is bigger as 
a patient tends to be overweight or obese.

We know from our simulation that sex does not influence the probability of 
missigness, nor is there any other factor that influences the probability of 
missingness, only the outcome itself. Therefore, conditioning on sex, will not 
eliminate the difference in distribution between observed and missing values.

```{r echo = FALSE}
ggplot(weights, aes(x = bmi)) +
  geom_histogram(bins = 50) +
  facet_grid(rows = vars(mnar), cols = vars(sex)) +
  theme_bw() +
  labs(
    title = "Distribution of BMI Conditional on Sex and Missingness (MNAR)",
    x = "Weight (kg)", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

Again, we see that, for each sex, the distribution of missing values is shifted
to the left compared to the distribution of observed values. This follows from
the fact that sex does not affect the probability of missingness. However, if we
condition on the BMI class (in other words, on the outcome itself), we see that
the difference between the distributions disappears.

```{r echo = FALSE}
ggplot(weights, aes(x = bmi)) +
  geom_histogram(bins = 50) +
  facet_grid(rows = vars(mnar), cols = vars(bmi_class)) +
  theme_bw() +
  labs(
    title = "Distribution of BMI Conditional on BMI and Missingness (MNAR)",
    x = "Weight (kg)", y = "Count"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
